---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

```{r}
knitr::opts_chunk$set(warning = FALSE)
knitr::opts_chunk$set(message = FALSE)
```


To read to whole *Peer Assessment 1* text click [here](https://github.com/itamaracampos/RepData_PeerAssessment1).
In this file is only show main points.

```{r}
#install.packages("tidyverse")
#install.packages("Hmisc")
```


```{r}
library(tidyverse) 
library(Hmisc)
library(lattice)
```


Note: the dataset and this file (plus code) must be in the same directory.

Save the plots in a directory. First create if don't exist.

```{r}
dir <- "figure"

if (!dir.exists(dir)){
dir.create(dir)
} else {
    print("Dir already exists!")
}
```




#### **1. Loading and preprocessing the data**

 **<span style="color: blue"> Solution  </span>**
 
Download the dataset from the cloud:

```{r}
 if(!file.exists("getdata-projectfiles-UCI HAR Dataset.zip")) {
        temp <- tempfile()
        download.file("http://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip",temp)
        unzip(temp)
        unlink(temp)
}
```


Or download the dataset from gihub directory, click in [View raw](https://github.com/itamaracampos/RepData_PeerAssessment1/blob/master/activity.zip).

```{r} 
unzip("./activity.zip")
activityData <- read.csv("./activity.csv")
activity <- read.csv("activity.csv")
```


```{r} 
head(activityData)
head(activity)

summary(activityData)
summary(activity)
```

 <span style="color: Green"> Download both ways show the same results. Chose one of those to continue. </span>

```{r} 
describe(activity)
```



#### **2. What is mean total number of steps taken per day?**

 **<span style="color: blue"> Solution </span>**

1. For this part of the assignment, ignore the missing values in the dataset. 1. Calculate the total number of steps taken per day.

```{r}
totalStepsByDay<-aggregate(steps~date, activity, na.rm = TRUE, sum )

```


2. Make a histogram of the total number of steps taken each day.ow same data, display using histogram. Depending how is display the plot can lead to wrong conclusion. 


```{r}
par(mfrow=c(1,2))

plot2a <- hist(totalStepsByDay$steps, xlab=" Total Number of Steps", ylab="Number of Days", 
     main="Number of Steps",
     col="lightblue1",
     border="dodgerblue3", ylim = c(0,30), labels=TRUE) 

plot2b <- hist(totalStepsByDay$steps, xlab=" Total Number of Steps", ylab="Number of Days", 
     main="Number of Steps",
     col="lightblue1",
     border="dodgerblue3", ylim = c(0,20),
     breaks = seq(0,30000, by=3000), labels=TRUE) #if you the test, play with this factor  breaks = seq(0,30000, by=3000)
```


```{r, echo=FALSE}

# Save the first picture above (on the left) if want use both just copy and past the previous code

File <- "./figure/Figure_Assessment1_task_02.png"
png(File)

plot2a <- hist(totalStepsByDay$steps, xlab=" Total Number of Steps", ylab="Number of Days", 
     main="Number of Steps",
     col="lightblue1",
     border="dodgerblue3", ylim = c(0,30), labels=TRUE) 


dev.off()

```

The **mean** of the total number of steps taken per day:

```{r}
mean(totalStepsByDay$steps)
```

**Median** of the total number of steps taken per day:

```{r}
median(totalStepsByDay$steps)
```




#### **3. What is the average daily activity pattern?**

 **<span style="color: blue"> Solution </span>**

1. Make a time series plot (i.e. type = “l”) of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

```{r}
average_daily_activity <- aggregate(activity$steps, by=list(activity$interval), FUN=mean, na.rm=TRUE)

names(average_daily_activity) <- c("interval", "mean")

plot(average_daily_activity$interval, average_daily_activity$mean, type = "l", 
     col="blue", lwd = 2, xlab="Interval", ylab="Average number of steps", main="Average number of steps per intervals")
```



```{r, echo=FALSE}

# Save the first picture above 
File <- "./figure/Figure_Assessment1_task_03.png"
png(File)

average_daily_activity <- aggregate(activity$steps, by=list(activity$interval), FUN=mean, na.rm=TRUE)

names(average_daily_activity) <- c("interval", "mean")

plot(average_daily_activity$interval, average_daily_activity$mean, type = "l", 
     col="blue", lwd = 2, xlab="Interval", ylab="Average number of steps", main="Average number of steps per intervals")

dev.off()

```

2. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

```{r}
average_daily_activity[which.max(average_daily_activity$mean), ]$interval
```





#### **4. Imputing missing values**

 **<span style="color: blue"> Solution </span>**

1. To remove the possibility of bias, we can remove the missing values. Which are:

```{r}
sum(is.na(activity$steps))
```

2. Devise a strategy for filling in all of the missing values in the dataset.


```{r}
all_steps <- average_daily_activity$mean[match(activity$interval, average_daily_activity$interval)]

glimpse(all_steps)
describe(all_steps)
```

3. Create a new dataset that is equal to the original dataset but with the missing data filled in

```{r}
activity_all <- transform(activity, steps = ifelse(is.na(activity$steps), yes = all_steps, no = activity$steps))

total_steps_all <- aggregate(steps ~ date, activity_all, sum)

names(total_steps_all) <- c("date", "daily_steps")

describe(total_steps_all)
```

4. Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day. The **mean** and **median** total number of steps taken per day.


```{r}
par(mfrow=c(1,2))

hist(total_steps_all$daily_steps, col = "pink", xlab = "Total steps per day", 
     main="Number of Steps",
     border="dodgerblue3", ylim = c(0,40), labels=TRUE) 

hist(total_steps_all$daily_steps, col = "pink", xlab = "Total steps per day", 
     main="Number of Steps",
     border="dodgerblue3", ylim = c(0,30),
     breaks = seq(0,30000, by=3000), labels=TRUE) #if you want to test, play with this factor  breaks = seq(0,30000, by=3000)


```



```{r, echo=FALSE}

# Save the first picture above on the left
File <- "./figure/Figure_Assessment1_task_04.png"
png(File)

hist(total_steps_all$daily_steps, col = "pink", xlab = "Total steps per day",
     main="Number of Steps",
     border="dodgerblue3", ylim = c(0,40), labels=TRUE) 

dev.off()

```

```{r}
mean(total_steps_all$daily_steps)
median(total_steps_all$daily_steps)
```



#### **5. Are there differences in activity patterns between weekdays and weekends?**


 **<span style="color: blue"> Solution </span>**

Use weekdays() function.Below two ways to plot the same data.

1. Create a new factor variable in the dataset with two levels -- "weekday" and "weekend"

```{r}
activity$date <- as.Date(strptime(activity$date, format="%Y-%m-%d"))

activity$datetype <- sapply(activity$date, function(x) {
  if (weekdays(x) == "Saturday" | weekdays(x) =="Sunday") 
        {y <- "Weekend"} else 
            {y <- "Weekday"}
            y
  })

describe(activity$datetype)
```

2. Make a panel plot containing a time series. Using i.e. type = "l. The plot should look something like the following, which was created using **simulated data**

```{r}
activity_by_date <- aggregate(steps~interval + datetype, activity, mean, na.rm = TRUE)

plot5a <- ggplot(activity_by_date, aes(x = interval , y = steps, color = datetype)) + geom_line() +
  labs(title = "Average daily steps by type of date", x = "Interval", y = "Average number of steps") +
   facet_wrap(~datetype, ncol = 1, nrow=2)
print(plot5a )
```


```{r}
plot5b <- with(activity_by_date, 
      xyplot(steps ~ interval | datetype, 
      type = "l",      
      main = "Total Number of Steps within Intervals by dayType",
      xlab = "Daily Intervals",
      ylab = "Average Number of Steps"))
print(plot5b )
```


```{r, echo=FALSE}

# Save both pictures above
File <- "./figure/Figure_Assessment1_task_05a.png"
png(File)
print(plot5a )
dev.off()

File <- "./figure/Figure_Assessment1_task_05b.png"
png(File)
print(plot5b )
dev.off()

```
